---
title: "quals_mds_GZD"
author: "Gwendolyn Donahue"
date: "2024-09-09"
output: html_document
---
# setup
## libraries
```{r}
library(tidyverse)
library(readxl)
library(cowplot)
library(ggplot2)
library(vegan)
library(lubridate)
library(RColorBrewer)
library(leaflet)
library(htmlwidgets)
```

## read in data
```{r}
qual_raw <- read_excel("/Users/gwendolyndonahue/Documents/GitHub/gwen-robin/quals/kfe_qual_data_entry_240823.xlsx")
```
## phylum assigning
```{r}
phylum_lookup <- tibble(
  name = c(
    # Porifera
    "Leucilla_nuttingi", "Leucosolenia_eleanor", "Craniella_arb", "Polymastia_pachymastia", 
    "Tethya_californiana", "Acanthancora_cyanocrypta",
    
    # Cnideria
    "Anthopleura_artemisia", "Anthopleura_sola", "Anthopleura_xanthogrammica", 
    "Cribrinopsis_albopunctata", "Urticina_piscivora", "Urticina_grebelneyi", 
    "Urticina_spp", "Metridium_farcimen", "Halcampa_decemtentaculata", 
    "Balanophyllia_elegans", "Paracyathus_stearnsi", "Corynactis_californica", 
    "Pachycerianthus_fimbriatus", "Virgulariidae_spp", "Stylaster_californicus", 
    "Aurelia_labiata", "Chrysaora_fuscesens", "Phacellophora_camtschatica",
    
    # Mollusca
    "Cryptochiton_stelleri", "Megathura_crenulata", "Pomaulax_gibberosa", 
    "Haliotis_kamtschatkana", "Haliotis_rufescens", "Neobernaya_spadicea", 
    "Ceratostoma_foliatum", "Kelletia_kelletii", "Cadlina_luteomarginata", 
    "Cadlina_flavomaculata", "Dialula_sandiegensis", "Rostanga_pulchra", 
    "Doriopsilla_albopunctata", "Doris_montereyensis", "Peltodoris_nobilis", 
    "Okenia_rosacea", "Triopha_catalinae", "Triopha_maculata", 
    "Aplysia_californica", "Hermissenda_opalescens", "Phidiana_hiltoni", 
    "Dendronotus_iris", "Chaceia_ovoidea", "Parapholas_californica", 
    "Crassodoma_gigantea", "Pododesmus_macrochisma", "Octopus_rubescens",
    
    # Annelida
    "Eudistylia_spp", "Myxicola_spp", "Serpula_columbiana", 
    "Dodecaceria_concharum", "Dodecaceria_fewkesi", "Diopatra_ornata", 
    "Salmacina_tribranchiata", "Phragmatopoma_californica",
    
    # Arthropoda
    "Balanus_nubilus", "Cancer_spp", "Loxorhynchus_crispatus", 
    "Loxorhynchus_grandis", "Pugettia_foliatus", "Pugettia_producta", 
    "Pugettia_richii", "Scyra_acutifrons",
    
    # Bryozoa
    "Diaperoforma_californica", "Celleporina_robertsoniae", "Celleporella_spp", 
    "Lagenicella_punctulata", "Hippoporina_insculpta", "Phidolopora_pacifica", 
    "Watersipora_subtorquata",
    
    # Brachiopoda
    "Terebretalia_transversa",
    
    # Echinodermata
    "Orthasterias_koehleri", "Leptasterias_spp", "Pisaster_brevispinus", 
    "Pisaster_giganteus", "Pycnopodia_helianthoides", "Dermasterias_imbricata", 
    "Mediaster_aequalis", "Patiria_miniata", "Henricia_spp", 
    "Amphiura_arcystata", "Ophioplocus_esmarki", "Ophiopterus_papillosa", 
    "Ophiothrix_spiculata", "Dendraster_excentricus", "Mesocentrotus_franciscanus", 
    "Strongylocentrotus_purpuratus", "Apostichopus_californicus", 
    "Apostichopus_parvimensis", "Cucumaria_miniata", "Cucumaria_piperata", 
    "Eupentacta_quinquesemita", "Psolus_chitonoides",
    
    # Tunicates
    "Boltenia_villosa", "Cnemidocarpa_finmarkiensis", "Halocynthia_igaboja", 
    "Pyura_haustor", "Styela_montereyensis", "Clavelina_huntsmani", 
    "Perophora_annectens", "Pycnoclavella_stanleyi", "Didemnum_carnulentum", 
    "Didemnum_orange_crust", "Aplidium_solidum", "Polyclinum_planum", 
    "Cystodytes_lobatus",
    
    # Vertebrates
    "Anarrhicthys_ocellatus", "Rhinogobiops_nicholsii", "Brachyistius_frenatus", 
    "Cymatogaster_aggregata", "Embiotoca_jacksoni", "Embiotoca_lateralis", 
    "Hypsurus_caryi", "Phanerodon_furcatus", "Rhacochilus_toxotes", 
    "Rhacochilus_vacca", "Oxyjulis_californica", "Semicossyphus_pulcher", 
    "Paralabrax_clathratus", "Girella_nigricans", "Medialuna_californiensis", 
    "Scorpaenichthys_marmoratus", "Sebastes_atrovirens", "Sebastes_auriculatus", 
    "Sebastes_carnatus", "Sebastes_caurinus", "Sebastes_chrysomelas", 
    "Sebastes_flavidus_serranoides", "Sebastes_maliger", "Sebastes_melanops", 
    "Sebastes_miniatus", "Sebastes_mystinus", "Sebastes_nebulosus", 
    "Sebastes_paucispinis", "Sebastes_pinniger", "Sebastes_serriceps", 
    "Hexagrammos_decagrammus", "Ophiodon_elongatus", "Oxylebius_pictus",
    
    # Brown algae
    "Stephanocystis_osmundacea", "Desmarestia_spp", "Alaria_marginata", 
    "Dictyoneurum_spp", "Egregia_menziesii", "Eisenia_arborea", 
    "Laminaria_setchellii", "Macrocystis_pyrifera", "Nereocystis_luetkeana", 
    "Pterygophora_californica",
    
    # Red algae
    "Cryptopleura_spp", "Bossiella_spp", "Calliarthron_spp", 
    "Callophyllis_spp", "Chondracanthus_corymbiferus", "Fredericqia_chiton", 
    "Sarcodiotheca_gaudichaudii", "Prionitis_spp", "Plocamium_spp", 
    "Botryocladia_pseudodichotoma", "Gloiocladia_laciniata", 
    "Fryeela_gardneri", "Rhodymenia_spp"
  ),
  phylum = c(
    # Porifera
    rep("Porifera", 6),
    # Cnideria
    rep("Cnideria", 18),
    # Mollusca
    rep("Mollusca", 27),
    # Annelida
    rep("Annelida", 8),
    # Arthropoda
    rep("Arthropoda", 8),
    # Bryozoa
    rep("Bryozoa", 7),
    # Brachiopoda
    rep("Brachiopoda", 1),
    # Echinodermata
    rep("Echinodermata", 22),
    # Tunicates
    rep("Tunicata", 13),
    # Vertebrates
    rep("Vertebrata", 33),
    # Brown algae
    rep("Brown_algae", 10),
    # Red algae
    rep("Red_algae", 13)
  )
)
```

## arrange sites north to south
```{r}
# rank CHNMS sites from north to south:
site_order <- tibble(
  site = c("Harmony Headlands 2", "Harmony Headlands 1", "Outer Buchon 1", 
           "Inner Buchon 2","Shearwater 1", "Shearwater 2", "Shearwater 4", 
           "White Rock 1", "White Rock 2","Outer Buchon 2", "Inner Buchon 1", 
           "Outer Buchon 3", "Jalama"),
  site_order_n_to_s = c(4, 3, 8, 6, 10, 12, 13, 1, 2, 9, 7, 5, 11))

all_site_order <- tibble(
  site = c("Shale Beds", "Monterey Wharf", "HMS90", "HMS100", "HMS Barren (Bird Rock)", "Otter Cove", "Carmel Pinnacles - Inner", "Carmel River State Beach", "Monastery-North", "Monastery-South", "Point Lobos", "White Rock 1", "White Rock 2", "Harmony Headlands 1", "Harmony Headlands 2", "Outer Buchon 3", "Inner Buchon 2", "Inner Buchon 1", "Outer Buchon 1", "Outer Buchon 2", "Shearwater 1", "Jalama", "Shearwater 2", "Shearwater 4"),
  all_site_order_up_to_down = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16, 17, 18, 19, 20, 21, 22, 23, 24)
)
```

# clean data
```{r}
qual_clean <- qual_raw %>% 
  # remove averaged rows so whole df is just rank abundance
  filter(!averaged_nums) %>% 
  # make 'NA' strings real NAs
  mutate(across(where(is.character), ~ na_if(.x, "NA"))) %>% 
  # (confirm this is acceptable...?): replace 'NI's with NAs
  mutate(across(where(is.character), ~ na_if(.x, "NI"))) %>% 
  # make necessary columns numeric instead of characters
  mutate(across(
    -c(date, site, region, observer, buddy, others_seen, notes_site, notes_other, averaged_nums),
    as.numeric)) %>% 
  select(-averaged_nums)
```
# subset for CHNMS sites
```{r}
qual_CHNMS <- qual_clean %>% 
  filter(site %in% c("Harmony Headlands 2", "Harmony Headlands 1", 
                     "Outer Buchon 1", "Inner Buchon 2", "Shearwater 1", 
                     "Shearwater 2", "Shearwater 4", "White Rock 1", 
                     "White Rock 2", "Outer Buchon 2", "Inner Buchon 1", 
                     "Outer Buchon 3", "Jalama"))

qual_CHNMS_limited <- qual_CHNMS %>% 
  mutate(year = year(date)) %>% 
  relocate(year, .before = site) %>% 
  select(-c(date, region, observer, buddy, depth_avg_ft, depth_max_ft,
            dive_duration_min, visibility_ft, bedrock, boulder, cobble, sand,
            others_seen, notes_site, notes_other, averaged_nums, site_num_mdsplot))
```
# CHNMS plots
## • MDS plots including all data
### - 1. set up mds clustering
```{r}
qual_CHNMS <- qual_CHNMS %>%
  mutate(site_num_mdsplot = row_number()) %>% 
  # create new substrate columns
  rowwise() %>%
  mutate(
    substrate_predominant = case_when(
      bedrock == 1 ~ "bedrock",
      boulder == 1 ~ "boulder",
      cobble == 1 ~ "cobble",
      sand == 1 ~ "sand",
      TRUE ~ NA_character_
    ),
    substrate_diversity = sum(!is.na(c_across(bedrock:sand)) & c_across(bedrock:sand) != 0)
  ) %>%
  ungroup() %>% 
  relocate(substrate_predominant, .after = sand) %>%
  relocate(substrate_diversity, .after = substrate_predominant)


qual_CHNMS_mds <- qual_CHNMS %>%
  # select species columns
  select(Leucilla_nuttingi:Rhodymenia_spp) %>%
  # CONSIDER CHANGING: replace all NAs (NIs) with 0
  mutate(across(where(is.numeric), ~ replace_na(., 0))) %>%
  as.matrix() %>%
  # measure distance for MDS using *bray
  # bray allows you to have lots of 0s
  # euclidean assumes some linear relationship (good for quantitative traits)
  vegdist(method = "bray") %>%
  metaMDS(k = 2)

# print stress value (as low as possible; needs to be under 0.2)
qual_CHNMS_mds$stress

# pull the MDS coords
mds_scores <- scores(qual_CHNMS_mds) %>%
  as_tibble(rownames = "site_num_mdsplot") %>%
  mutate(site_num_mdsplot = as.integer(site_num_mdsplot)) %>%
  # pull back in columns of interest
  left_join(qual_CHNMS %>% select(site_num_mdsplot, 
                                  site, 
                                  depth_max_ft, visibility_ft,
                                  substrate_predominant, substrate_diversity), 
            by = "site_num_mdsplot")
```
### - mds by site
```{r}
# plot MDS
ggplot(mds_scores, aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(color = site)) +
  theme_minimal() +
  labs(title = "NMDS Plot of CHNMS dives (2023-2024)", x = "NMDS Dimension 1", y = "NMDS Dimension 2")
```
### - mds by grouped sites
```{r}
# all HH grouped; all IB grouped; all OB grouped, etc.
```

### - mds by point divisions
```{r}
# add column to designate which sites are north of point buchon
mds_scores_point_designations <- mds_scores %>%
  mutate(north_buchon = case_when(
    site == "Harmony Headlands 2" ~ TRUE,
    site == "Harmony Headlands 1" ~ TRUE,
    site == "Outer Buchon 1" ~ FALSE,
    site == "Inner Buchon 2" ~ FALSE,
    site == "Shearwater 1" ~ FALSE,
    site == "Shearwater 2" ~ FALSE,
    site == "Shearwater 4" ~ FALSE,
    site == "White Rock 1" ~ TRUE,
    site == "White Rock 2" ~ TRUE,
    site == "Outer Buchon 2" ~ FALSE,
    site == "Inner Buchon 1" ~ FALSE,
    site == "Outer Buchon 3" ~ TRUE,
    site == "Jalama" ~ FALSE)) %>% 
  mutate(north_conception = case_when(
    site == "Harmony Headlands 2" ~ TRUE,
    site == "Harmony Headlands 1" ~ TRUE,
    site == "Outer Buchon 1" ~ TRUE,
    site == "Inner Buchon 2" ~ TRUE,
    site == "Shearwater 1" ~ TRUE,
    site == "Shearwater 2" ~ FALSE,
    site == "Shearwater 4" ~ FALSE,
    site == "White Rock 1" ~ TRUE,
    site == "White Rock 2" ~ TRUE,
    site == "Outer Buchon 2" ~ TRUE,
    site == "Inner Buchon 1" ~ TRUE,
    site == "Outer Buchon 3" ~ TRUE,
    site == "Jalama" ~ TRUE))

# plot MDS divided by point buchon
ggplot(mds_scores_point_designations, aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(color = north_buchon)) +
  theme_minimal() +
  labs(title = "NMDS Plot of CHNMS sites '23-'24 (divided by point buchon)", x = "NMDS Dimension 1", y = "NMDS Dimension 2")
# plot MDS divided by point conception
ggplot(mds_scores_point_designations, aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(color = north_conception)) +
  theme_minimal() +
  labs(title = "NMDS Plot of CHNMS sites '23-'24 (divided by point conception)", x = "NMDS Dimension 1", y = "NMDS Dimension 2")
```
### - mds  by depth
```{r}
ggplot(mds_scores, aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(color = depth_max_ft)) +
  scale_color_gradient(low = "lightsteelblue1", high = "#141c44") +
  theme_minimal() +
  labs(title = "NMDS Plot of CHNMS dives (2023-2024)", x = "NMDS Dimension 1", y = "NMDS Dimension 2")
```

### - mds by substrate diversity
```{r}
ggplot(mds_scores, aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(color = substrate_diversity)) +
  scale_color_gradient(low = "pink", high = "#141c44") +
  theme_minimal() +
  labs(title = "NMDS Plot of CHNMS dives (2023-2024)", x = "NMDS Dimension 1", y = "NMDS Dimension 2")
```
### - mds plot by visibility
```{r}
ggplot(mds_scores, aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(color = visibility_ft)) +
  scale_color_gradient(low = "#D2B48C", high = "#003152") +
  theme_minimal() +
  labs(title = "NMDS Plot of CHNMS dives (2023-2024)", x = "NMDS Dimension 1", y = "NMDS Dimension 2")
```
## • MDS plots (scorched earth version)
### - 1. set up mds clustering
```{r}
# scorch (remove where necessary because of NIs)
qual_scorchedCHNMS <- qual_CHNMS %>%
  # remove white rock 2
  filter(site != "White Rock 2") 

# save columns to add back in before removing 
save_scorched_columns <- qual_scorchedCHNMS %>%
  select(site, 
         depth_max_ft, 
         visibility_ft,
         substrate_predominant, 
         substrate_diversity) %>%
  mutate(site_num_scorched_mdsplot = row_number())

# now remove all columns not going into the matrix
qual_scorchedCHNMS <- qual_scorchedCHNMS %>% 
  # select all species columns
  select(Leucilla_nuttingi:Rhodymenia_spp) %>%
  # remove columns where there were NIs we don't feel comfortable subbing 0 in
  select(-c(Pomaulax_gibberosa, 
            #crabs
            Cancer_spp, Loxorhynchus_crispatus,
            Loxorhynchus_grandis, Pugettia_foliatus, Pugettia_producta,
            Pugettia_richii, Scyra_acutifrons, 
            #bryozoans
            Diaperoforma_californica,Celleporina_robertsoniae,
            Celleporella_spp,Lagenicella_punctulata,Hippoporina_insculpta,
            Phidolopora_pacifica,Watersipora_subtorquata,
            #psolus
            Psolus_chitonoides,
            #red algaes
            Cryptopleura_spp,Bossiella_spp,Calliarthron_spp,
            Callophyllis_spp,Chondracanthus_corymbiferus,Fredericqia_chiton,
            Sarcodiotheca_gaudichaudii,Prionitis_spp,Plocamium_spp,
            Botryocladia_pseudodichotoma,Gloiocladia_laciniata,
            Fryeela_gardneri,Rhodymenia_spp,
            #polyclinum
            Polyclinum_planum,Acanthancora_cyanocrypta, Urticina_grebelneyi, Corynactis_californica, Cadlina_flavomaculata, Rostanga_pulchra, Peltodoris_nobilis, Triopha_maculata, Phidiana_hiltoni, Pododesmus_macrochisma, Serpula_columbiana, Salmacina_tribranchiata, Phragmatopoma_californica, Terebretalia_transversa, Boltenia_villosa, Cnemidocarpa_finmarkiensis, Halocynthia_igaboja, Pyura_haustor, Styela_montereyensis, Clavelina_huntsmani, Perophora_annectens, Pycnoclavella_stanleyi, Didemnum_carnulentum,Didemnum_orange_crust)) %>% 
  # replace remaining NAs (NIs) with 0
    # other NAs:
      # Acanthancora_cyanocrypta, Urticina_grebelneyi, Corynactis_californica, Cadlina_flavomaculata, Rostanga_pulchra, Peltodoris_nobilis, Triopha_maculata, Phidiana_hiltoni, Pododesmus_macrochisma, Serpula_columbiana, Salmacina_tribranchiata, Phragmatopoma_californica, Terebretalia_transversa, Boltenia_villosa, Cnemidocarpa_finmarkiensis, Halocynthia_igaboja, Pyura_haustor, Styela_montereyensis, Clavelina_huntsmani, Perophora_annectens, Pycnoclavella_stanleyi, Didemnum_carnulentum,Didemnum_orange_crust
  mutate(across(where(is.numeric), ~ replace_na(., 0)))

# create MDS matrix
qual_scorchedCHNMS_mds <- qual_scorchedCHNMS %>% 
  as.matrix() %>%
  vegdist(method = "bray") %>%
  metaMDS(k = 2)

# print stress value (as low as possible; needs to be under 0.2)
qual_scorchedCHNMS_mds$stress

# pull the MDS coords
scorched_mds_scores <- scores(qual_scorchedCHNMS_mds) %>%
  as_tibble(rownames = "site_num_scorched_mdsplot") %>%
  mutate(site_num_scorched_mdsplot = as.integer(site_num_scorched_mdsplot)) %>%
  # pull back in columns of interest
  left_join(save_scorched_columns %>% select(
    site_num_scorched_mdsplot, 
    site, 
    depth_max_ft, visibility_ft,
    substrate_diversity), 
    by = "site_num_scorched_mdsplot")
```
### - mds by site
```{r}
ggplot(scorched_mds_scores, aes(x = NMDS1, y = NMDS2)) +
  geom_point(aes(color = site)) +
  theme_minimal() +
  labs(title = "scorched NMDS Plot of CHNMS dives (2023-2024)", x = "NMDS Dimension 1", y = "NMDS Dimension 2")
```

### - *TO DO
```{r}
# colored by geography (rough area) 
# different shaped points to show if site is in an MPA or not
```

## • Abundance plots
### - 1. set up data
```{r}
# set up long df (each row is one species at one dive)
qual_CHNMS_abundances <- qual_CHNMS_limited %>% 
  pivot_longer(cols = 
                 starts_with("Leucilla_nuttingi"):
                 ends_with("Rhodymenia_spp"),
               names_to = "species",
               values_to = "abundance") %>% 
  # make site a factored column so we can arrange N to S
  left_join(site_order, by = "site") %>%
  mutate(site = factor(site, 
                       levels = site_order$site[
                         order(site_order$site_order_n_to_s)])) %>% 
  # add phylum column
  left_join(phylum_lookup, by = c("species" = "name"))

# filter to 2024
qual_CHNMS_abundances_2024 <- qual_CHNMS_abundances %>% 
  filter(year == 2024)
```
### - basic abundance plot
```{r}
ggplot(qual_CHNMS_abundances_2024, aes(x = site, y = abundance, fill = species)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Species Abundance by Sites (2024)",
       x = "Site (arranged N to S)",
       y = "Species Abundance") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "none")
```

### - basic phyla abundance plot
```{r}
ggplot(qual_CHNMS_abundances_2024, aes(x = site, y = abundance, fill = phylum)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(title = "Phyla Abundance by Sites (2024)",
       x = "Site (arranged N to S)",
       y = "Phylum Abundance") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),
        legend.position = "right")
```

## • Evenness-Richness plot
### - prep data
```{r}
site_metrics_2024 <- qual_CHNMS_abundances_2024 %>%
  filter(abundance > 0) %>%
  group_by(site) %>%
  summarize(
    # calculate species richness (# unique species observed)
    richness = n_distinct(species), 
    # calculate species diversity 
    shannon_diversity = -sum(
      (abundance / sum(abundance)) * log(abundance / sum(abundance) + 1e-10), 
      na.rm = TRUE), 
    # calculate evenness (species diversity normalized by richness)
    evenness = shannon_diversity / log(richness + 1e-10))
```

### - plot
```{r}
ggplot(site_metrics_2024, 
       aes(x = richness, y = evenness)) +
  geom_point(aes(color = site), size = 3) +
  labs(
    title = "2024 CHNMS Overview of Sites (qual sheets)",
    x = "Species Richness",
    y = "Evenness"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text.y = element_text(size = 8)
  )
```
## • Richness (# sp per phylum)
### 1. set up data
```{r}
# prep
CHNMS_2024_site_phyla <- qual_CHNMS_abundances_2024 %>%
  select(site, phylum) %>%
  distinct() %>%
  complete(site, phylum)

# calc richness (total number of species per phylum observed per site)
temp_CHNMS_2024_richness_df <- qual_CHNMS_abundances_2024 %>%
  # use presence/absence of each observation
  filter(abundance > 0) %>%
  group_by(site, phylum) %>%
  # find num unique species per site and phylum
  summarize(richness = n_distinct(species), .groups = 'drop')

# combine to keep 0s if that phylum wasn't observed at that site
CHNMS_2024_richness <- CHNMS_2024_site_phyla %>%
  left_join(temp_CHNMS_2024_richness_df, by = c("site", "phylum")) %>%
  replace_na(list(richness = 0)) %>% 
  filter(site != "White Rock 2") %>% #too many NIs
  # no qual sheets in 2024:
  filter(site != "Shearwater 1") %>% 
  filter(site != "Shearwater 2") %>% 
  # consider leaving in? but we saw none
  filter(phylum != "Brachiopoda")
```

### - heatmap plot
```{r}
ggplot(CHNMS_2024_richness, 
       aes(x = site, y = phylum, fill = richness)) +
  geom_tile() +
  scale_fill_gradient(low = "white", 
                      high = "red", 
                      na.value = "lightgrey", # specify color for NAs
                      name = "# Species per Phylum") +
  theme_minimal() +
  labs(title = "Species Richness Heatmap (2024)", 
       x = "Site (N to S)", 
       y = "Phylum") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  theme(axis.text.y = element_text(size = 8))
```
### - stacked bar plot
```{r}
# create color palette (11 colors)
colors_phyla <- brewer.pal(11, "Paired")

# plot
ggplot(CHNMS_2024_richness, 
       aes(x = site, y = richness, fill = phylum)) +
  geom_bar(stat = "identity", position = "stack") +
  #get rid of gap between bars and x axis
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)))+
  labs(title = "Species Richness (2024)",
       x = "Site (arranged N to S)",
       y = "# Species Present") +
  scale_fill_manual(values = colors_phyla) +
  theme_half_open() +
  background_grid() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "right",
        # make grid lines thinner
        panel.grid.major = element_line(color = "grey80", linewidth = 0.2), 
        panel.grid.minor = element_line(color = "grey90", linewidth = 0.2))
```
## • TO DO: Spatial turnover plot
```{r}

```
## • TO DO: kelp vs urchins (look at HH and Buchon)
# All qual sheets (both CHNMS and Monterey and Carmel)
## • set up data
### - general cleaning
```{r}
qual_clean <- qual_clean %>% 
  select(-c(depth_avg_ft, others_seen, notes_site, notes_other)) %>% 
  rowwise() %>%
  mutate(
    # fill this new column with whichever substrate was ranked '1'
    substrate_predominant = case_when(
      bedrock == 1 ~ "bedrock",
      boulder == 1 ~ "boulder",
      cobble == 1 ~ "cobble",
      sand == 1 ~ "sand",
      TRUE ~ NA_character_),
    # fill this new column with the number of substrates given a number
    substrate_diversity = sum(!is.na(c_across(bedrock:sand)) & 
                                c_across(bedrock:sand) != 0)) %>%
  ungroup() %>% 
  relocate(substrate_predominant, .after = sand) %>%
  relocate(substrate_diversity, .after = substrate_predominant)

# consider changing
# average qual sheets if the sites are the same
qual_clean_averaged <- qual_clean %>% 
  group_by(site) %>%
  summarise(
    across(where(is.numeric), ~ round(mean(., na.rm = TRUE))),
    # keep first instances of character columns (region and main substrate)
    region = first(region),
    substrate_predominant = first(substrate_predominant),
    .groups = "drop")
```
### - nmds matrix setup
```{r}

```
### - richness (presence / absence) setup
```{r}
# first make long df
qual_abundances <- qual_clean_averaged %>% 
  pivot_longer(cols = 
                 starts_with("Leucilla_nuttingi"):
                 ends_with("Rhodymenia_spp"),
               names_to = "species",
               values_to = "abundance") %>% 
  # make site a factored column so we can arrange N to S
  left_join(all_site_order, by = "site") %>%
  mutate(site = factor(site, 
                       levels = all_site_order$site[
                         order(all_site_order$all_site_order_up_to_down)])) %>% 
  # add phylum column
  left_join(phylum_lookup, by = c("species" = "name")) %>% 
  select(-c(bedrock, boulder, cobble, sand))

# next find species richness
# 1. prep
qual_site_phyla <- qual_abundances %>%
  select(site, phylum) %>%
  distinct() %>%
  complete(site, phylum)
# 2. calc richness (total number of species per phylum observed per site)
temp_qual_richness_df <- qual_abundances %>%
  filter(abundance > 0) %>%
  group_by(site, phylum) %>%
  summarize(richness = n_distinct(species), .groups = 'drop')
# 3. combine to keep 0s if that phylum wasn't observed at that site
qual_richness <- qual_site_phyla %>%
  left_join(temp_qual_richness_df, by = c("site", "phylum")) %>%
  replace_na(list(richness = 0))
```
## • MDS plots
```{r}

```

## • Species richness plots
```{r}
# create color palette (11 colors)
colors_12_phyla <- brewer.pal(12, "Paired")

# plot
ggplot(qual_richness, 
       aes(x = site, y = richness, fill = phylum)) +
  geom_bar(stat = "identity", position = "stack") +
  #get rid of gap between bars and x axis
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)))+
  labs(title = "Species Richness (all quals)",
       x = "Site (arranged upcoast to downcoast)",
       y = "# Species Present") +
  scale_fill_manual(values = colors_12_phyla) +
  theme_half_open() +
  background_grid() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        legend.position = "right",
        # make grid lines thinner
        panel.grid.major = element_line(color = "grey80", linewidth = 0.2), 
        panel.grid.minor = element_line(color = "grey90", linewidth = 0.2))
```
# Map plots
##• set up data (site level data)
```{r}
# read in coord data
qual_coords <- read_excel("/Users/gwendolyndonahue/Documents/GitHub/gwen-robin/quals/qual_coords_estimate_240911.xlsx")

qual_coords <- qual_coords %>% 
  # make numeric not characters
  mutate(across(c(latitude, longitude),as.numeric))

qual_coords_norepeats <- qual_coords %>% 
  # keep only one coord per site (right now just keeping first since they're the same)
  group_by(site) %>%
  summarise(latitude = first(latitude),
            longitude = first(longitude))

# merge with working df
qual_richness_coords <- qual_richness %>% 
  left_join(qual_coords_norepeats %>% select(site, latitude, longitude,), 
            by = "site")

# richness summary df
# add columns for # phyla present, most speciose phylum at that site?
site_richness_summary <- qual_richness_coords %>% 
  group_by(site, latitude, longitude) %>%
  summarize(total_richness = sum(richness, na.rm = TRUE))

# add in other site info of interest
site_final_summary <- site_richness_summary %>% 
  
```
## • all qual sheets
```{r}
# create richness color palette
colors_richness <- colorNumeric(
  palette = "Spectral", 
  domain = site_richness_summary$total_richness)

# plot
leaflet(data = site_richness_summary) %>%
  addTiles() %>%  
  addCircleMarkers(~longitude, ~latitude, 
                   color = ~colors_richness(total_richness), 
                   fillOpacity = 1, radius = 2,
                   popup = ~paste(site, "species richness = ", total_richness)) %>%
  addLegend(pal = colors_richness, values = ~site_richness_summary$total_richness, 
            title = "Species richness",
            position = "topright") %>%
  addScaleBar(position = 'bottomleft')
```
## • CHNMS map
```{r}
# pull in coords for non-diving sites too
chnms_site_notes <- read_excel("/Users/gwendolyndonahue/Documents/GitHub/gwen-robin/quals/CHNMS_Site_Notes.xlsx")

chnms_site_notes_filtered <- chnms_site_notes %>% 
  # temporarily sub in white rock coords
  mutate(Latitude_2023 = ifelse(Site_name %in% c("White_Rock_1", "White_Rock_2"), 
                                Latitude_2024, Latitude_2023),
         Longitude_2023 = ifelse(Site_name %in% c("White_Rock_1", "White_Rock_2"), 
                                 Longitude_2024, Longitude_2023)) %>% 
  # select columns of interest
  select(c("Site_name", 
           "Dive_site",
           "Latitude_2023", "Longitude_2023", 
           "Protected_area_2024", 
           "Depth_avg", 
           "2024_data_collected"))
```

```{r}
# leaflet note: can't change shape directly
# --> means you can either have different shapes or colors but not both

# Subset data
chnms_protected_areas <- chnms_site_notes_filtered %>% filter(Protected_area_2024 == TRUE)
chnms_non_protected_areas <- chnms_site_notes_filtered %>% filter(Protected_area_2024 == FALSE)

getColor <- function(Dive_site) {
  ifelse(Dive_site == TRUE, "#08306b", "black")}

leaflet() %>%
  addTiles() %>%
  
  # protected sites: solid circle in the middle of point
  addCircleMarkers(
    data = chnms_protected_areas,
    lng = ~as.numeric(Longitude_2023), lat = ~as.numeric(Latitude_2023), 
    fillOpacity = 2, radius = 2, # Slightly larger to differentiate
    color = ~getColor(Dive_site),
    popup = ~paste0("<b>", Site_name, #bolded
                    "</b><br/>",#line break
                    "Data collected in Fall August 2024: ", `2024_data_collected`)) %>%
  
  # non-protected sites: no solid circle in the middle of point
  addCircleMarkers(
    data = chnms_non_protected_areas,
    lng = ~as.numeric(Longitude_2023), lat = ~as.numeric(Latitude_2023), 
    fillOpacity = 0.01, radius = 2, 
    color = ~getColor(Dive_site),
    popup = ~paste0("<b>", Site_name,
                    "</b><br/>",
                    "Data collected in Fall August 2024: ", `2024_data_collected`)) %>%
  
  addLegend(
    position = "topright",     
    colors = c("#08306b", "black"), labels = c("Dive Site", "Non-Dive Site"), 
    title = htmltools::HTML(
    "<strong>Site Type</strong>
    <br/>
    <span style='font-size:10px;
    '>Note: Solid center indicates site is in a protected area </span>"), 
    opacity = 1
  ) %>% 
  
  addScaleBar(position = 'bottomleft')
```

# NOTES:
## possible to-dos
- make figures for fraction (or %) observed instead of species richness